Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-03-25 manuscript.tex
latest updates

The integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
ldeuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution is calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
on the value of the integrals is used.

@@

The summation over Bessel function indices in Eq. \ref{eq:dielectrictensor}
continues until the relative change between the current and
previous iteration is less than a tolerance of $10^{-8}$. Root finding of Eq.
\ref{eq:dettensor} in complex frequency plane is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method,
whereby if the values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes when the real extent of
the simplex is less than $10^{-4}\, \Omega_i$ and the imaginary extent is less
than $10^{-5}\,\Omega_i$, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than the Nelder-Mead method or winding number approach used in isolation
of one another. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a laptop.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
The observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates. Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes
cease to grow once the fastest growing mode becomes
nonlinear\cite{Cook2013}. Hence, the $log$ of a mode's amplitude is
proportional to its growth rate under the assumption that
that all modes start growing in unison from identical seed amplitudes.
There are several points worth noting about this figure. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

This effect can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Other possible mechanisms for doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

>>>>

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution are calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.

@@

Bessel functions indices in Eq. \ref{eq:dielectrictensor} are summed outwards
from 0 in both the positive and negative directions
until the relative change between the current and
previous iteration is less than a tolerance of $10^{-8}$. Root finding of Eq.
\ref{eq:dettensor} in complex frequency plane is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method,
whereby if the values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes when the real extent of
the simplex is less than $10^{-4}\, \Omega_i$ and the imaginary extent is less
than $10^{-5}\,\Omega_i$, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than the Nelder-Mead method or winding number approach used in isolation
of one another. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a laptop.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
The observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates.
Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes rising
as $\exp(\gamma t)$, where $\gamma$ is a mode's growth rate,
cease to grow once the fastest growing mode becomes
nonlinear\cite{Cook2013}.
Therefore under the assumption that all modes start growing in unison from
identical seed amplitudes then the log of a mode's amplitude is
proportional to its growth rate.
There are several points worth noting about this figure. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Other possible mechanisms for doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.
